---
title: "Sample Demographics"
subtitle: "Demographic and Clinical Characteristics by Group (Table 1)"
author: "Jeffrey Girard"
date: "2026-10-07"
format:
  html:
    code-fold: true
    toc: true
---

::: {.callout-note}
This report reads participant-level demographic and clinical records that are
not publicly shared, so it cannot be re-rendered from this repository. Only
aggregate results are shown. The model covariates (age, sex, race, and PTSD
status) are included in the shared feature data.
:::

## Setup

```{r}
#| message: false
#| label: setup

library(tidyverse)
library(gt)
```

### Data

```{r}
# Restricted REDCap exports (not shared)
redcap_dir <- Sys.getenv("CMBA_REDCAP_DIR", r"(Z:\CMBA_PTSD\1_RAWDATA\redCAP)")

redcap_vars <- c(
  "record_id", "race", "ethnicity", "sex", "age", "income", "education",
  "phq_total", "pcl5_m_total", "caps_mnth_total", "caps_mnth_ptsd",
  "suds_account", "suds_impact"
)

read_redcap <- function(file, group) {
  read_csv(
    file.path(redcap_dir, file),
    col_types = cols(.default = col_character())
  ) |>
    select(all_of(redcap_vars)) |>
    summarize(
      .by = record_id,
      group = group,
      across(everything(), \(x) first(x, na_rm = TRUE))
    )
}

demo_all <-
  bind_rows(
    read_redcap("CMBA-Patients_2022-12-05_1438.csv", "clinic"),
    read_redcap("CMBA-Control_2023-03-10_1522.csv", "prolific")
  ) |>
  # participants who completed the behavioral tasks and diagnostic interview
  filter(!is.na(suds_account), !is.na(caps_mnth_ptsd)) |>
  # participant 199 completed the session, but their recording was never
  # segmented into tasks, so they have no behavioral features and are not
  # analyzed
  filter(record_id != "199") |>
  transmute(
    record_id,
    GRP = factor(group),
    Race = factor(
      as.numeric(race),
      levels = 1:7,
      labels = c(
        "American Indian",
        "Asian",
        "Black",
        "Hispanic",
        "More than one race",
        "White",
        "Unknown"
      )
    ),
    Ethnicity = factor(
      as.numeric(ethnicity),
      levels = 1:3,
      labels = c(
        "Not Hispanic or Latino",
        "Hispanic or Latino",
        "Unknown"
      )
    ),
    Sex = factor(
      as.numeric(sex),
      levels = 1:6,
      labels = c(
        "Male",
        "Female",
        "Intersex",
        "Transgender (sex assigned at birth female, now male)",
        "Transgender (sex assigned at birth male, now female)",
        "Prefer not to say"
      )
    ),
    Income = factor(
      as.numeric(income),
      levels = 1:7,
      labels = c(
        "< $10,000",
        "$10,000-19,999",
        "$20,000-39,999",
        "$40,000-79,999",
        "$80,000-169,999",
        "$170,000+",
        "Prefer not to say"
      )
    ),
    Age = as.numeric(age),
    Education = as.numeric(education),
    PHQ9 = parse_number(phq_total),
    PCL5 = parse_number(pcl5_m_total),
    CAPS5 = parse_number(caps_mnth_total),
    PTSD = factor(caps_mnth_ptsd, levels = 1:0, labels = c("PTSD", "Control")),
    SUDS_Account = parse_number(suds_account),
    SUDS_Impact = parse_number(suds_impact)
  ) |>
  # one participant selected "Hispanic" as their race; they are reported as
  # such here and coded as nonwhite in the analyses
  mutate(across(Race:Sex, \(x) fct_na_value_to_level(x, "Unknown")))

stopifnot(nrow(demo_all) == 92)

```

### Sample

```{r}
n_group <- table(demo_all$PTSD)

demo_all |>
  count(GRP, PTSD) |>
  pivot_wider(names_from = PTSD, values_from = n, values_fill = 0) |>
  mutate(
    GRP = str_to_title(GRP),
    Total = PTSD + Control
  ) |>
  gt(rowname_col = "GRP") |>
  tab_stubhead(label = "Recruitment source") |>
  tab_spanner(label = "Diagnostic group (CAPS-5)", columns = c(PTSD, Control)) |>
  grand_summary_rows(columns = c(PTSD, Control, Total), fns = list(Total ~ sum(.)), fmt = ~ fmt_integer(.)) |>
  cols_align(align = "center", columns = c(PTSD, Control, Total))
```

## Group Comparisons

```{r}
# Formatting helpers (APA style: no leading zero for p and bounded effect sizes)
fmt_p <- function(p) {
  if_else(p < .001, "< .001", sub("^0", "", sprintf("%.3f", p)))
}
fmt_es <- function(x) {
  sub("^(-?)0", "\\1", sprintf("%.2f", x)) |> sub(pattern = "^-", replacement = "−")
}

# Mann-Whitney U test with rank-biserial correlation (positive = higher in PTSD group)
mwu_test <- function(data, var) {
  data <- filter(data, !is.na(.data[[var]]))
  x <- data[[var]][data$PTSD == "PTSD"]
  y <- data[[var]][data$PTSD == "Control"]
  w <- wilcox.test(x, y, exact = FALSE)
  tibble(
    test = "Mann-Whitney U",
    statistic = sprintf("U = %.1f", w$statistic),
    p = w$p.value,
    es_name = "r",
    es = unname(2 * w$statistic / (length(x) * length(y)) - 1),
    n = length(x) + length(y)
  )
}

# Chi-square or Fisher's exact test with Cramér's V
nominal_test <- function(data, var, exact) {
  tab <- table(droplevels(data[[var]]), data$PTSD)
  chi <- suppressWarnings(chisq.test(tab, correct = FALSE))
  v <- sqrt(unname(chi$statistic) / (sum(tab) * (min(dim(tab)) - 1)))
  result <- if (exact) {
    tibble(test = "Fisher's exact", statistic = "", p = fisher.test(tab)$p.value)
  } else {
    tibble(
      test = "Chi-square",
      statistic = sprintf("χ²(%d) = %.2f", chi$parameter, chi$statistic),
      p = chi$p.value
    )
  }
  mutate(result, es_name = "V", es = v, n = sum(tab), min_expected = min(chi$expected))
}

comparisons <- bind_rows(
  # Expected counts are all >= 5 for sex
  nominal_test(demo_all, "Sex", exact = FALSE) |> mutate(variable = "Sex"),
  # Some expected counts are < 5 for race and ethnicity
  nominal_test(demo_all, "Race", exact = TRUE) |> mutate(variable = "Race"),
  nominal_test(filter(demo_all, Ethnicity != "Unknown"), "Ethnicity", exact = TRUE) |>
    mutate(variable = "Ethnicity"),
  # Income is ordinal, so compare the ordered income brackets
  demo_all |>
    filter(Income != "Prefer not to say") |>
    mutate(Income_rank = as.integer(droplevels(Income))) |>
    mwu_test("Income_rank") |>
    mutate(variable = "Income"),
  mwu_test(demo_all, "Age") |> mutate(variable = "Age (years)"),
  mwu_test(demo_all, "Education") |> mutate(variable = "Education (years)"),
  mwu_test(demo_all, "SUDS_Account") |> mutate(variable = "SUDS, Account"),
  mwu_test(demo_all, "SUDS_Impact") |> mutate(variable = "SUDS, Impact"),
  mwu_test(demo_all, "PCL5") |> mutate(variable = "PCL-5")
  # CAPS-5 is not tested because the groups were defined by the CAPS-5 diagnosis
) |>
  relocate(variable)

comparisons |>
  gt(rowname_col = "variable") |>
  fmt(columns = es, fns = fmt_es) |>
  fmt_number(columns = min_expected, decimals = 1) |>
  fmt(columns = p, fns = fmt_p) |>
  sub_missing(missing_text = "") |>
  cols_merge(columns = c(es_name, es), pattern = "{1} = {2}") |>
  cols_label(
    test = "Test",
    statistic = "Statistic",
    p = md("*p*"),
    es_name = "Effect size",
    n = md("*n*"),
    min_expected = "Min. expected count"
  ) |>
  cols_align(align = "center", columns = -c(variable, test)) |>
  tab_source_note(md(paste(
    sprintf(
      "Ethnicity excludes \"Unknown\" (n = %d) and income excludes \"Prefer not to say\" (n = %d).",
      sum(demo_all$Ethnicity == "Unknown"), sum(demo_all$Income == "Prefer not to say")
    ),
    "*V* = Cramér's *V*; *r* = rank-biserial correlation, positive when values are higher in the PTSD group.",
    "Income was compared as ordered brackets.",
    "CAPS-5 was not tested because it defined the groups."
  )))
```

Because the income test excludes participants who declined to report income,
we also compared the rate of declining by group:

```{r}
income_nonresponse <- table(
  `Declined to report income` = demo_all$Income == "Prefer not to say",
  demo_all$PTSD
)

income_nonresponse |>
  as.data.frame.matrix() |>
  rownames_to_column("Declined to report income") |>
  mutate(`Declined to report income` = if_else(`Declined to report income` == "TRUE", "Yes", "No")) |>
  gt() |>
  cols_align(align = "center", columns = c(PTSD, Control)) |>
  tab_source_note(sprintf(
    "Fisher's exact test, p = %s",
    fmt_p(fisher.test(income_nonresponse)$p.value)
  ))
```

## Table 1

```{r}
tabulate_var <- function(var) {
  tab <- table(droplevels(demo_all[[var]]), demo_all$PTSD)
  pct <- 100 * prop.table(tab, 2)
  tibble(
    row = rownames(tab),
    c1 = as.character(tab[, "PTSD"]),
    c2 = sprintf("%.1f%%", pct[, "PTSD"]),
    c3 = as.character(tab[, "Control"]),
    c4 = sprintf("%.1f%%", pct[, "Control"]),
    type = "level"
  )
}

describe_var <- function(var, label) {
  x <- na.omit(demo_all[[var]][demo_all$PTSD == "PTSD"])
  y <- na.omit(demo_all[[var]][demo_all$PTSD == "Control"])
  tibble(
    row = label,
    c1 = sprintf("%.2f", mean(x)), c2 = sprintf("%.2f", sd(x)),
    c3 = sprintf("%.2f", mean(y)), c4 = sprintf("%.2f", sd(y)),
    type = "continuous"
  )
}

test_cols <- comparisons |>
  transmute(row = variable, p = fmt_p(p), es = paste0("*", es_name, "* = ", fmt_es(es)))

continuous_vars <- c(
  "Age" = "Age (years)",
  "Education" = "Education (years)",
  "SUDS_Account" = "SUDS, Account",
  "SUDS_Impact" = "SUDS, Impact",
  "PCL5" = "PCL-5",
  "CAPS5" = "CAPS-5"
)

table_1 <- bind_rows(
  tibble(row = "Categorical variables", c1 = "n", c2 = "%", c3 = "n", c4 = "%", type = "header"),
  map(c("Sex", "Race", "Ethnicity", "Income"), \(v) {
    bind_rows(tibble(row = v, type = "variable"), tabulate_var(v))
  }) |> list_rbind(),
  tibble(row = "Continuous variables", c1 = "M", c2 = "SD", c3 = "M", c4 = "SD", type = "header"),
  imap(continuous_vars, \(label, var) describe_var(var, label)) |> list_rbind()
) |>
  left_join(test_cols, by = "row") |>
  mutate(
    across(c(p, es), \(x) if_else(row == "CAPS-5", "—", x)),
    across(c(c1:c4, p, es), \(x) replace_na(x, ""))
  )

table_1 |>
  gt() |>
  cols_hide(type) |>
  tab_header(
    title = "Table 1",
    subtitle = "Demographic and Clinical Characteristics by Group"
  ) |>
  tab_spanner(label = sprintf("PTSD (n = %d)", n_group[["PTSD"]]), columns = c(c1, c2)) |>
  tab_spanner(label = sprintf("Control (n = %d)", n_group[["Control"]]), columns = c(c3, c4)) |>
  cols_label(row = "", c1 = "", c2 = "", c3 = "", c4 = "", p = md("*p*"), es = "Effect size") |>
  fmt_markdown(columns = es) |>
  cols_width(row ~ px(210), c(c1, c3) ~ px(55), c(c2, c4) ~ px(75), p ~ px(60), es ~ px(85)) |>
  # keep Quarto from overriding the column widths
  tab_options(quarto.disable_processing = TRUE) |>
  cols_align(align = "center", columns = c(c1:c4, p, es)) |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_body(rows = type == "header")) |>
  tab_style(style = cell_text(style = "italic"), locations = cells_body(columns = c1:c4, rows = type == "header")) |>
  tab_style(style = cell_text(indent = px(12)), locations = cells_body(columns = row, rows = type %in% c("variable", "continuous"))) |>
  tab_style(style = cell_text(indent = px(28)), locations = cells_body(columns = row, rows = type == "level")) |>
  tab_source_note(md(paste(
    sprintf(
      "*Note.* The control group includes %d participants recruited through Prolific and %d recruited from the clinic who did not meet criteria for PTSD.",
      sum(demo_all$GRP == "prolific" & demo_all$PTSD == "Control"),
      sum(demo_all$GRP == "clinic" & demo_all$PTSD == "Control")
    ),
    "Percentages are column percentages within each group.",
    if (any(demo_all$Race == "Hispanic")) "One participant selected Hispanic as their race (coded as nonwhite in the analyses)." else "",
    "Sex was compared with a chi-square test, and race and ethnicity with Fisher's exact test",
    "because some expected cell counts were below 5; effect sizes are Cramér's *V*.",
    "Income (as ordered brackets) and continuous variables were compared with Mann-Whitney *U* tests;",
    "effect sizes are rank-biserial correlations (*r*), positive when values are higher in the PTSD group.",
    "Tests of ethnicity and income excluded \"Unknown\" and \"Prefer not to say\" responses, respectively.",
    "CAPS-5 was not tested because it defined the groups.",
    "SUDS = Subjective Units of Distress Scale, rated after the trauma account and impact statement tasks;",
    "PCL-5 = PTSD Checklist for DSM-5; CAPS-5 = Clinician-Administered PTSD Scale for DSM-5 total severity."
  )))
```
